Fast Sorting on Modern Computers

### 2005-07-31 docs/tech_report/heap.tex
more minor renaming

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. Cache heapsort also
has a similar increase, though LaMarca's results show that these occur at the
same position in the graph, in other words, when the set size is the same. Our
results, however, show that this occurs earlier in memory-tuned heapsort, due to the
fact that it is an out of place sort.

>>>>

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. Memory-tuned
heapsort also has a similar increase, though LaMarca's results show that these
occur at the same position in the graph, in other words, when the set size is
the same. Our results, however, show that this occurs earlier in memory-tuned
heapsort, due to the fact that it is an out of place sort.

### 2005-07-31 docs/tech_report/merge.tex
more minor renaming

\section{Double Multi-mergesort}

Double Multi-mergesort combines the double tiling of Double Tiled mergesort with
Multi-mergesort.

@@

As predicted, the rate of level 1 misses is greatly reduced by the double
tiling, as shown in Figure \ref{merge level 1 misses}. Double multi-mergesort
has consistently four times fewer level 1 misses than multi-mergesort, and both
double tiled sorts perform significantly better than the unoptimized sorts. The
expected increase in level 2 misses did not materialise, and is in fact
negligible.

@@

The cache misses end up being the deciding factor between the performance of
double-aligned tiled mergesort and base mergesort, the next best contender.
Despite the reduction in level 2 cache misses, the sharp increase in level 1
misses result is poor performance for tiled and multi-mergesort. Multi-mergesort
is further encumbered with a much higher instruction count. Double tiled
mergesort, with its low instruction count, and low cache miss rate in both
levels 1 and 2, emerges the winner.

@@

Double-tiled mergesort (which has the same results are tiled mergesort) has a
predilection for merging forwards. This is likely to be due to the fact that the
final merge goes forwards. The left merge step is also favoured, but this is
merely due to the fact that it occurs first in the code. While there are more
left branches, the actual number of branches taken are the same.

>>>>

\section{Double-aligned Multi-mergesort}

Double-aligned Multi-mergesort combines the double alignment of Double-aligned
tiled mergesort with multi-mergesort.

@@

As predicted, the rate of level 1 misses is greatly reduced by the double
tiling, as shown in Figure \ref{merge level 1 misses}. Double-aligned multi-mergesort
has consistently four times fewer level 1 misses than multi-mergesort, and both
double tiled sorts perform significantly better than the unoptimized sorts. The
expected increase in level 2 misses did not materialise, and is in fact
negligible.

@@

The cache misses end up being the deciding factor between the performance of
double-aligned tiled mergesort and base mergesort, the next best contender.
Despite the reduction in level 2 cache misses, the sharp increase in level 1
misses result is poor performance for tiled and multi-mergesort. Multi-mergesort
is further encumbered with a much higher instruction count. Double-aligned tiled
mergesort, with its low instruction count, and low cache miss rate in both
levels 1 and 2, emerges the winner.

@@

Double-aligned tiled mergesort (which has the same results are tiled mergesort) has a
predilection for merging forwards. This is likely to be due to the fact that the
final merge goes forwards. The left merge step is also favoured, but this is
merely due to the fact that it occurs first in the code. While there are more
left branches, the actual number of branches taken are the same.

### 2005-07-31 docs/tech_report/radix.tex
minor edits
more minor renaming

Cache radixsort should traverse the array only nine times: Once at the start to
count the least significant bytes; six times (once per array per pass) as it
copies using the three least significant bytes (counting the next least
significant byte at the same time), and a final two as it copies using the most
significant byte.

\subsection{Aligned Memory-tuned Radixsort}
In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
with reference to it auxiliary array. Specifically, extra memory is allocated
for the auxiliary array, which is then offset so that the start of the two
arrays are exactly half the length of the cache apart. We further added an
alignment for the level 1 cache (see Section \ref{double-aligned tiled
mergesort}) to significantly reduce level 1 misses at the expense of a small
number of level 2 misses. We adapt this strategy here, and call it Aligned Cache
Radixsort.

@@

Due to the reduced copying, memory-tuned radixsort should have a lower instruction
count and fewer cache misses that base radixsort. There should be nine passes
over the array, leading to just over one miss per key, as opposed to two misses
per key with base radixsort. The complexity and branch prediction results
should remain the same. The instruction count should, however, reduce by nearly
a half to $fract{9}{16}$ths of base radixsort.

@@

Cache radixsort performs as predicted as well. By sorting keys between two
arrays instead of copying a partially sorted array back, the instruction count
has been reduced by more than 10\%, and the number of level 2 cache misses
have been halved, to one and a quarter miss per key. This is exactly
as expected, due to there being nine passes over the array, and a one miss
per pass per cache line (coming to slightly under one and a quarter. The extra
is again probably due to the unpredictable order of the copying).

>>>>

Memory-tuned radixsort should traverse the array only nine times: Once at the
start to count the least significant bytes; six times (once per array per pass)
as it copies using the three least significant bytes (counting the next least
significant byte at the same time), and a final two as it copies using the most
significant byte.

\subsection{Aligned Memory-tuned Radixsort}
In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
with reference to it auxiliary array. Specifically, extra memory is allocated
for the auxiliary array, which is then offset so that the start of the two
arrays are exactly half the length of the cache apart. We further added an
alignment for the level 1 cache (see Section \ref{double-aligned tiled
mergesort}) to significantly reduce level 1 misses at the expense of a small
number of level 2 misses. We adapt this strategy here, and call it Aligned
Memory-tuned Radixsort.

@@

Due to the reduced copying, memory-tuned radixsort should have a lower instruction
count and fewer cache misses that base radixsort. There should be nine passes
over the array, leading to just over one miss per key, as opposed to two misses
per key with base radixsort. The complexity and branch prediction results
should remain the same. The instruction count should, however, reduce by nearly
a half to $\frac{9}{16}^ths$ of base radixsort.

@@

Memory-tuned radixsort performs as predicted as well. By sorting keys between two
arrays instead of copying a partially sorted array back, the instruction count
has been reduced by more than 10\%, and the number of level 2 cache misses
have been halved, to one and a quarter miss per key. This is exactly
as expected, due to there being nine passes over the array, and a one miss
per pass per cache line (coming to slightly under one and a quarter. The extra
is again probably due to the unpredictable order of the copying).
